---
title: "MSI MALDI Portfolio Workflow"
author: ""
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
format:
  html:
    toc: true
    toc_depth: 3
    code-fold: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This document outlines the complete workflow for processing and
analyzing a MALDI mass‑spectrometry imaging (MSI) dataset.

The overall goal is to load raw MSI data, perform a series of
preprocessing steps (normalisation, peak picking, alignment and
filtering), inspect and visualise the data, optionally crop a region of
interest (ROI), carry out spatial segmentation and principal component
analysis, and finish with optional downstream analyses. Each major step
is encapsulated in a separate script and sourced in this Quarto document
for clarity and reproducibility.

## Data and project structure

The data directory contains the following files:

```         
data
├── MB001-040_pos.tif          # Optical scan of the tissue section
├── N-glycan_RScript.docx       # Original analysis notes
├── README.txt                  # Dataset description
├── cal-nonormalization.ibd     # Binary data for the calibrant region
├── cal-nonormalization.imzML   # Metadata for the calibrant region
├── control-nonormalization.ibd # Binary data for the control region
├── control-nonormalization.imzML
├── png1-nonormalization.ibd    # Binary data for treated region 1
├── png1-nonormalization.imzML
├── png2-nonormalization.ibd    # Binary data for treated region 2
└── png2-nonormalization.imzML
```

## Workflow overview

Each of the following sections sources an R script from the `scripts/`
directory. The scripts are created separately and contain the code
required to execute each step. The Quarto document executes the scripts
sequentially when rendered. If you modify script file names, update the
paths accordingly. Tables and figures will also be saved in `tables/`
and `figures/`, respectively.

# Data loading

Load the raw imzML and ibd files into a Cardinal `MSImageSet` object.

```{r load-data, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
source("scripts/01_load_data.R")
```

```{r load-data-summary, echo=FALSE}
if (exists("msi_dataset_overview", inherits = TRUE)) {
  knitr::kable(
    msi_dataset_overview,
    caption = "Loaded MSI datasets and spectral coverage."
  )
}
```

# Preprocessing

Perform total‑ion current (TIC) normalization, peak picking, alignment
and filtering. Adjust thresholds and tolerances within the script as
required.

```{r preprocess-data, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
source("scripts/02_preprocess_data.R")
```

```{r preprocessing-summary, echo=FALSE, eval=TRUE}
if (exists("msi_preprocessing_summary", inherits = TRUE)) {
  prep_tbl <- transform(
    msi_preprocessing_summary,
    feature_delta = raw_features - processed_features,
    pixel_delta = raw_pixels - processed_pixels
  )
  knitr::kable(
    prep_tbl,
    caption = "Effect of preprocessing on pixel/feature counts."
  )
}
```

# Quality control and exploration

Generate diagnostic plots, ion images and spectrum views to assess data
quality both before and after preprocessing.

```{r explore-data, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
source("scripts/03_explore_data.R")
```

```{r qc-top-features, echo=FALSE}
if (exists("msi_qc_spatial_features", inherits = TRUE) &&
    length(msi_qc_spatial_features) > 0) {
  spatial_tbl <- tibble::enframe(msi_qc_spatial_features, name = "dataset", value = "mz_values")
  spatial_tbl$top_spatial_mz <- purrr::map_chr(spatial_tbl$mz_values, ~ paste(.x, collapse = ", "))
  spatial_tbl$mz_values <- NULL
  knitr::kable(spatial_tbl, caption = "Top spatially structured m/z values per dataset.")
}
```

```{r qc-figures, echo=FALSE, results='asis'}
qc_legend <- list(
  tic = "*Figure legend:* Raw vs. processed TIC heatmaps; processed images should look smoother and more uniform.",
  spectra = "*Figure legend:* Raw vs. processed spectra from a representative pixel; peak picking/alignment simplify the processed trace.",
  spatial = "*Figure legend:* Top spatially structured ions; bright regions highlight m/z values that localize to specific tissue areas."
)
if (!exists("msi_qc_plots", inherits = TRUE) ||
    length(msi_qc_plots) == 0) {
  qc_dir <- file.path("figures", "qc")
  if (dir.exists(qc_dir)) {
    qc_files <- list.files(
      qc_dir,
      pattern = "_qc_(tic|spectra|spatial)\\.png$",
      full.names = TRUE
    )
    if (length(qc_files) > 0) {
      qc_df <- tibble::tibble(
        file = qc_files,
        dataset = gsub("_qc_.*$", "", basename(qc_files)),
        panel = gsub("^.*_qc_(tic|spectra|spatial)\\.png$", "\\1", basename(qc_files))
      )
      qc_split <- split(qc_df, qc_df$dataset)
      msi_qc_plots <- purrr::map(qc_split, function(df) {
        list(
          tic = df$file[df$panel == "tic"][1],
          spectra = df$file[df$panel == "spectra"][1],
          spatial = df$file[df$panel == "spatial"][1]
        )
      })
    }
  }
}
if (exists("msi_qc_plots", inherits = TRUE) && length(msi_qc_plots) > 0) {
  render_img <- function(path) {
    if (!is.null(path) && file.exists(path)) {
      cat(sprintf("![](%s)\n\n", knitr::image_uri(path)))
      return(TRUE)
    }
    return(FALSE)
  }

  purrr::iwalk(msi_qc_plots, function(paths, dataset_name) {
    cat(sprintf("### QC Report for %s\n\n", dataset_name))
    
    tic_shown <- render_img(paths$tic)
    if (tic_shown) cat(qc_legend$tic, "\n\n")
    spectra_shown <- render_img(paths$spectra)
    if (spectra_shown) cat(qc_legend$spectra, "\n\n")
    spatial_shown <- render_img(paths$spatial)
    if (spatial_shown) cat(qc_legend$spatial, "\n\n")

    # Fallback for legacy single-panel QC images if per-panel files are missing.
    if (!tic_shown && !spectra_shown && !spatial_shown) {
      legacy_path <- file.path("figures", "qc", paste0(dataset_name, "_qc.png"))
      if (render_img(legacy_path)) {
        cat("\n")
      }
    }
  })
} else {
  cat("*(QC figures will appear after generating them in the quality control chunk.)*")
}
```

# ROI selection and cropping

Optionally crop the dataset to a region of interest. This step is useful
for speeding up exploration or focusing on a particular anatomical area.
If you prefer to analyse the full dataset, you may omit this section.

```{r roi-crop, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
source("scripts/04_roi_crop.R")
```

```{r roi-summary, echo=FALSE}
if (exists("msi_roi_summary", inherits = TRUE)) {
  knitr::kable(
    msi_roi_summary,
    caption = "Pixel counts before and after ROI cropping."
  )
}
```

# Segmentation: Identifying Molecular Regions

This step partitions the image into distinct molecular regions based on
spectral similarity. We will apply and compare three different
clustering methods to understand the benefits of spatially-aware
algorithms.

The script `05_ssc_segmentation.R` performs two main tasks:

1.  **Optimizing `k`**: It first runs the Spatial Shrunken Centroids
    (SSC) algorithm for a range of `k` values (number of segments),
    plotting each result to help us choose a `k` that best fits the
    data's complexity.

2.  **Comparing Methods**: For a fixed `k`, it generates a side-by-side
    comparison of three methods:

    -   **Spatial Shrunken Centroids (SSC)**: A method that 
    incorporates spatial smoothing and performs feature
        selection.

    -   **Spatial K-Means (SKM)**: A faster spatial method that adds a
        smoothing penalty to the standard k-means algorithm.

    -   **K-Means (non-spatial)**: Standard k-means clustering performed
        on the PCA scores, used here as a baseline.

```{r segmentation, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
source("scripts/05_ssc_segmentation.R")
```

```{r segmentation-summary, echo=FALSE}
if (exists("msi_segmentation_summary", inherits = TRUE) &&
    nrow(msi_segmentation_summary) > 0) {
  knitr::kable(
    msi_segmentation_summary,
    caption = "Segmentation sweep parameters and comparison coverage."
  )
}
```

```{r segmentation-figures, echo=FALSE, results='asis'}
emit_seg_fig <- function(title, path) {
  if (is.null(path) || !file.exists(path)) return()
  img_uri <- knitr::image_uri(path)
  cat(sprintf("#### %s\n\n![](%s)\n\n", title, img_uri))
}
seg_legends <- list(
  sweep = "*Figure legend:* SSC segmentation grids across k values; each panel shows spatial clusters, with higher k revealing finer subregions.",
  compare = "*Figure legend:* Method comparison at fixed k; SSC and SKM should yield smoother contiguous regions than non-spatial k-means."
)
if (!exists("msi_segmentation_plots", inherits = TRUE) ||
    length(msi_segmentation_plots) == 0) {
  seg_dir <- file.path("figures", "segmentation")
  if (dir.exists(seg_dir)) {
    seg_files <- list.files(seg_dir, pattern = "\\.png$", full.names = TRUE)
    if (length(seg_files) > 0) {
      seg_names <- tools::file_path_sans_ext(basename(seg_files))
      msi_segmentation_plots <- purrr::set_names(seg_files, seg_names)
    }
  }
}
if (exists("msi_segmentation_plots", inherits = TRUE) &&
    length(msi_segmentation_plots) > 0) {
  sweep_paths <- msi_segmentation_plots[grepl("_ssc_grid$", names(msi_segmentation_plots))]
  comparison_paths <- msi_segmentation_plots[grepl("_comparison$", names(msi_segmentation_plots))]
  if (length(sweep_paths) > 0) {
    cat("### SSC k-sweep panels\n\n")
    purrr::iwalk(sweep_paths, function(path, label) emit_seg_fig(label, path))
    cat(seg_legends$sweep, "\n\n")
  }
  if (length(comparison_paths) > 0) {
    cat("### Method comparison panels\n\n")
    purrr::iwalk(comparison_paths, function(path, label) emit_seg_fig(label, path))
    cat(seg_legends$compare, "\n\n")
  }
} else {
  cat("*(Segmentation figures are available after running the segmentation chunk.)*")
}
```

### Interpreting the Results

The code chunk above will produce two sets of plots:

1.  **SSC Results (Varying k)**: This plot shows the tissue segmentation
    for each `k` value tested. By visually inspecting these images, you
    can choose a `k` that appears to best capture the underlying
    anatomical structures without over-segmenting the tissue into noisy,
    meaningless regions.

2.  **Segmentation Method Comparison**: This 3-panel plot directly
    contrasts the output of SSC, SKM, and non-spatial K-Means. You will
    likely observe that the K-Means result has a noisy,
    "salt-and-pepper" appearance. In contrast, the SKM and SSC results
    should appear much smoother and more contiguous, as these algorithms
    penalize solutions where neighboring pixels are assigned to 
    different clusters. This demonstrates the power of including 
    spatial information in the clustering process.

# Principal component analysis (PCA)

Perform dimensionality reduction (e.g. PCA) to extract dominant spectral
patterns and visualize spatial loadings.

```{r pca-analysis, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
source("scripts/06_pca_analysis.R")
```

```{r pca-figures, echo=FALSE, results='asis'}
emit_pca_fig <- function(title, path) {
  if (is.null(path) || !file.exists(path)) return()
  cat(sprintf("#### %s\n\n![](%s)\n\n", title, knitr::image_uri(path)))
}
pca_legends <- list(
  scores = "*Figure legend:* PCA score maps show spatial patterns captured by each principal component; smooth gradients indicate dominant anatomical variation.",
  loadings = "*Figure legend:* PCA loading spectra highlight m/z values driving each component; labeled peaks are candidate biomarkers."
)
if (!exists("msi_pca_score_paths", inherits = TRUE) ||
    length(msi_pca_score_paths) == 0 ||
    !all(file.exists(unlist(msi_pca_score_paths)))) {
  pca_dir <- file.path("figures", "pca")
  if (dir.exists(pca_dir)) {
    score_files <- list.files(pca_dir, pattern = "_scores\\.png$", full.names = TRUE)
    load_files <- list.files(pca_dir, pattern = "_loadings\\.png$", full.names = TRUE)
    if (length(score_files) > 0) {
      msi_pca_score_paths <- purrr::set_names(score_files, tools::file_path_sans_ext(basename(score_files)))
    }
    if (length(load_files) > 0) {
      msi_pca_loading_paths <- purrr::set_names(load_files, tools::file_path_sans_ext(basename(load_files)))
    }
  }
}
if (exists("msi_pca_score_paths", inherits = TRUE) &&
    length(msi_pca_score_paths) > 0) {
  cat("### PCA score maps\n\n")
  purrr::iwalk(msi_pca_score_paths, function(path, label) emit_pca_fig(label, path))
  cat(pca_legends$scores, "\n\n")
}
if (exists("msi_pca_loading_paths", inherits = TRUE) &&
    length(msi_pca_loading_paths) > 0) {
  cat("### PCA loading spectra\n\n")
  purrr::iwalk(msi_pca_loading_paths, function(path, label) emit_pca_fig(label, path))
  cat(pca_legends$loadings, "\n\n")
} else {
  cat("*(PCA figure panels will display after running the PCA analysis chunk.)*")
}
```

```{r pca-summary, echo=FALSE}
if (exists("msi_pca_loading_summary", inherits = TRUE) &&
    nrow(msi_pca_loading_summary) > 0) {
  knitr::kable(
    msi_pca_loading_summary,
    caption = "Representative m/z drivers per principal component."
  )
}
```

# Additional analyses

Include any extra analyses not covered above, such as t‑SNE/UMAP,
supervised classification, co‑localisation networks or annotation-based
exploration. Create additional scripts as required and source them here.

```{r extra-analysis, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
source("scripts/07_extra_analysis.R")
```

```{r extra-analysis-output, echo=FALSE, results='asis'}
emit_extra_fig <- function(title, path) {
  if (is.null(path) || !file.exists(path)) return()
  cat(sprintf("#### %s\n\n![](%s)\n\n", title, knitr::image_uri(path)))
}
extra_legends <- list(
  manifold = "*Figure legend:* t-SNE/UMAP embeddings map pixel spectra into 2D; tight, well-separated clusters imply distinct spectral profiles aligned with segmentation labels.",
  network = "*Figure legend:* Co-localization networks connect m/z features that co-occur across pixels; thicker edges indicate stronger co-occurrence."
)
if (!exists("msi_manifold_paths", inherits = TRUE) ||
    length(msi_manifold_paths) == 0) {
  extra_dir <- file.path("figures", "extra")
  if (dir.exists(extra_dir)) {
    tsne_files <- list.files(extra_dir, pattern = "_tsne\\.png$", full.names = TRUE)
    umap_files <- list.files(extra_dir, pattern = "_umap\\.png$", full.names = TRUE)
    all_ds <- unique(c(
      gsub("_tsne\\.png$", "", basename(tsne_files)),
      gsub("_umap\\.png$", "", basename(umap_files))
    ))
    if (length(all_ds) > 0) {
      msi_manifold_paths <- purrr::map(purrr::set_names(all_ds), function(ds) {
        list(
          tsne = tsne_files[basename(tsne_files) == paste0(ds, "_tsne.png")][1],
          umap = umap_files[basename(umap_files) == paste0(ds, "_umap.png")][1]
        )
      })
    }
  }
}
if (!exists("msi_network_paths", inherits = TRUE) ||
    length(msi_network_paths) == 0) {
  extra_dir <- file.path("figures", "extra")
  if (dir.exists(extra_dir)) {
    net_files <- list.files(extra_dir, pattern = "_network\\.png$", full.names = TRUE)
    if (length(net_files) > 0) {
      net_names <- gsub("_network\\.png$", "", basename(net_files))
      msi_network_paths <- purrr::set_names(net_files, net_names)
    }
  }
}
if (exists("msi_manifold_paths", inherits = TRUE) &&
    length(msi_manifold_paths) > 0) {
  cat("### Manifold learning embeddings\n\n")
  purrr::iwalk(msi_manifold_paths, function(paths, dataset) {
    if (length(paths) == 0) return()
    cat("#### ", dataset, "\n\n")
    purrr::iwalk(paths, function(path, method) emit_extra_fig(method, path))
  })
  cat(extra_legends$manifold, "\n\n")
}

if (exists("msi_classifier_summary", inherits = TRUE) &&
    nrow(msi_classifier_summary) > 0) {
  knitr::kable(
    msi_classifier_summary,
    caption = "Random forest accuracy and exported importance plots."
  )
}

if (exists("msi_network_paths", inherits = TRUE) &&
    length(msi_network_paths) > 0) {
  cat("### Co-localization networks\n\n")
  purrr::iwalk(msi_network_paths, function(path, dataset) {
    emit_extra_fig(dataset, path)
  })
  cat(extra_legends$network, "\n\n")
} else {
  cat("*(Network diagrams will appear once the extra analysis chunk finishes running.)*")
}
```

# Feature annotation

Use the significant m/z values from PCA loadings and classifier
importance to seed downstream database searches (METLIN, HMDB, GNPS).

```{r annotate-features, echo=TRUE, message=TRUE, warning=FALSE, eval=FALSE}
source("scripts/08_annotate_features.R")
```

```{r annotation-summary, echo=FALSE, eval=FALSE}
if (exists("msi_annotation_preview", inherits = TRUE)) {
  knitr::kable(
    head(msi_annotation_preview, 20),
    caption = "First 20 m/z values queued for annotation (full list saved to tables/significant_mz_features.csv)."
  )
}
```

# Conclusions

Summarise your findings from each analysis step and discuss any
biological interpretations or follow‑up questions. This section can also
document challenges encountered and potential improvements for future
analyses.

```{r conclusions-summary, echo=FALSE, eval=FALSE}
if (exists("msi_preprocessing_summary", inherits = TRUE)) {
  total_raw_pixels <- sum(msi_preprocessing_summary$raw_pixels, na.rm = TRUE)
  total_proc_pixels <- sum(msi_preprocessing_summary$processed_pixels, na.rm = TRUE)
  total_features_removed <- sum(msi_preprocessing_summary$feature_delta, na.rm = TRUE)
  cat(
    sprintf(
      "* Preprocessing retained %s/%s pixels while removing %s redundant features across the cohort.\n",
      format(total_proc_pixels, big.mark = ","),
      format(total_raw_pixels, big.mark = ","),
      format(total_features_removed, big.mark = ",")
    )
  )
}
if (exists("msi_segmentation_summary", inherits = TRUE) &&
    nrow(msi_segmentation_summary) > 0) {
  k_val <- unique(na.omit(msi_segmentation_summary$comparison_k))[1]
  cat(sprintf("* Segmentation compared SSC, SKM, and K-means at k = %s; review the method grids to confirm spatial smoothness.\n", k_val))
}
if (exists("msi_classifier_summary", inherits = TRUE) &&
    nrow(msi_classifier_summary) > 0) {
  best_acc <- max(msi_classifier_summary$test_accuracy, na.rm = TRUE)
  cat(sprintf("* Random forest classifiers reached %.1f%% held-out accuracy, validating that spectral differences align with spatial regions.\n", best_acc))
}
if (exists("msi_annotation_preview", inherits = TRUE)) {
  cat("* Exported `tables/significant_mz_features.csv` lists high-priority ions for METLIN/HMDB lookups and future MS/MS confirmation.\n")
}
```
