---
title: "MSI MALDI Portfolio Workflow"
author: ""
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
format:
  html:
    toc: true
    toc_depth: 3
  pdf:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Introduction

This document outlines the complete workflow for processing and analyzing a
MALDI mass‑spectrometry imaging (MSI) dataset.  

The overall goal is to load raw MSI data, perform a series of preprocessing
steps (normalisation, peak picking, alignment and filtering), inspect and
visualise the data, optionally crop a region of interest (ROI), carry out
spatial segmentation and principal component analysis, and finish with
optional downstream analyses.  Each major step is encapsulated in a separate
script and sourced in this Quarto document for clarity and reproducibility.

## Data and project structure

The data directory contains the following files:

```
data
├── MB001-040_pos.tif          # Optical scan of the tissue section
├── N-glycan_RScript.docx       # Original analysis notes
├── README.txt                  # Dataset description
├── cal-nonormalization.ibd     # Binary data for the calibrant region
├── cal-nonormalization.imzML   # Metadata for the calibrant region
├── control-nonormalization.ibd # Binary data for the control region
├── control-nonormalization.imzML
├── png1-nonormalization.ibd    # Binary data for treated region 1
├── png1-nonormalization.imzML
├── png2-nonormalization.ibd    # Binary data for treated region 2
└── png2-nonormalization.imzML
```

## Workflow overview

Each of the following sections sources an R script from the `scripts/`
directory.  The scripts are created separately and contain the
code required to execute each step.  The Quarto document executes
the scripts sequentially when rendered.  If you modify script file names, 
update the paths accordingly.

# Data loading

Load the raw imzML and ibd files into a Cardinal `MSImageSet` object.

```{r load-data, echo=TRUE, message=FALSE, warning=FALSE}
source("scripts/01_load_data.R")
```

# Preprocessing

Perform total‑ion current (TIC) normalization, peak picking, alignment and
filtering. Adjust thresholds and tolerances within the script as required.

```{r preprocess-data, echo=TRUE, message=FALSE, warning=FALSE}
source("scripts/02_preprocess_data.R")
```

# Quality control and exploration

Generate diagnostic plots, ion images and spectrum views to
assess data quality both before and after preprocessing.

```{r explore-data, echo=TRUE, message=FALSE, warning=FALSE}
source("scripts/03_explore_data.R")
```

# ROI selection and cropping

Optionally crop the dataset to a region of interest.  This step is
useful for speeding up exploration or focusing on a particular
anatomical area.  If you prefer to analyse the full dataset, you may
omit this section.

```{r roi-crop, echo=TRUE, message=FALSE, warning=FALSE}
source("scripts/04_roi_crop.R")
```

# Segmentation: Identifying Molecular Regions

This step partitions the image into distinct molecular regions based on
spectral similarity. We will apply and compare three different clustering
methods to understand the benefits of spatially-aware algorithms.

The script `05_ssc_segmentation.R` now performs two main tasks:
1.  **Optimizing `k`**: It first runs the Spatial Shrunken Centroids (SSC)
    algorithm for a range of `k` values (number of segments), plotting each
    result to help us choose a `k` that best fits the data's complexity.
2.  **Comparing Methods**: For a fixed `k`, it generates a side-by-side
    comparison of three methods:
    - **Spatial Shrunken Centroids (SSC)**: A sophisticated method that
      incorporates spatial smoothing and performs feature selection.
    - **Spatial K-Means (SKM)**: A faster spatial method that adds a
      smoothing penalty to the standard k-means algorithm.
    - **K-Means (non-spatial)**: Standard k-means clustering performed on
      the PCA scores, used here as a baseline.

```{r segmentation, echo=TRUE, message=FALSE, warning=FALSE}
source("scripts/05_ssc_segmentation.R")
```

### Interpreting the Results

The code chunk above will produce two sets of plots:

1.  **SSC Results (Varying k)**: This plot shows the tissue segmentation for
    each `k` value tested. By visually inspecting these images, you can choose
    a `k` that appears to best capture the underlying anatomical structures
    without over-segmenting the tissue into noisy, meaningless regions.

2.  **Segmentation Method Comparison**: This 3-panel plot directly contrasts
    the output of SSC, SKM, and non-spatial K-Means. You will likely observe
    that the K-Means result has a noisy, "salt-and-pepper" appearance. In
    contrast, the SKM and SSC results should appear much smoother and more
    contiguous, as these algorithms penalize solutions where neighboring
-   pixels are assigned to different clusters. This demonstrates the power
    of including spatial information in the clustering process.


# Principal component analysis (PCA)

Perform dimensionality reduction (e.g. PCA) to extract dominant
spectral patterns and visualize spatial loadings.

```{r pca-analysis, echo=TRUE, message=FALSE, warning=FALSE}
source("scripts/06_pca_analysis.R")
```

# Additional analyses

Include any extra analyses not covered above, such as t‑SNE/UMAP,
supervised classification, co‑localisation networks or
annotation-based exploration.  Create additional scripts as required
and source them here.

```{r extra-analysis, echo=TRUE, message=FALSE, warning=FALSE}
source("scripts/07_extra_analysis.R")
```

# Conclusions

Summarise your findings from each analysis step and discuss any
biological interpretations or follow‑up questions.  This section can
also document challenges encountered and potential improvements for
future analyses.
